---
title: "spatialTV Package Demo: Spatial Total Variation for Deconvolution"
author: "Tom Liu"
date: "`r Sys.Date()`"
output: html_document
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

The **spatialTV** package implements spatial Total Variation (TV) regularization for spatial transcriptomics deconvolution. Given initial cell type estimates from methods like RCTD, spatialTV refines these estimates by enforcing spatial coherence while learning a low-rank mixture representation.

## Method Overview

The package implements a three-step workflow:

1. **Input $S$**: Cell type likelihood scores from deconvolution (e.g., RCTD)
2. **Build spatial graph $A$**: Delaunay triangulation of spatial coordinates
3. **Optimize**: Learn mixture proportions $X$ and dictionary $B$ with TV regularization

The objective function combines a negative log-likelihood term with a weighted Total Variation penalty:
$$\min_{X,B} -\sum_{i=1}^N \log z_i + \lambda \sum_{e=1}^m w_e \sum_{d=1}^D \sqrt{(AX)_{ed}^2 + \delta}$$

where $z_i = \sum_k S_{ik}(XB^\top)_{ik}$ and $X$ represents spot-level mixture proportions while $B$ is the cell type dictionary.

## Installation

Install the spatialTV package from GitHub:

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("liutom445/spatialTV")
```

Load required packages:

```{r libraries}
suppressPackageStartupMessages({
  library(spatialTV)
  library(spacexr)      # For RCTD
  library(Matrix)
  library(ggplot2)
  library(gridExtra)    # For multi-panel plots
  library(tidyr)        # For data reshaping
})
```

# From RCTD to $S$

We extract $S$ from `myRCTD@results$singlet_scores` and transform scores with a softmax-like temperature (`SCALE`) so rows remain positive (row sums not enforced).

```{r}
myRCTD <- readRDS("myRCTD_08_updated.rds")
n <- myRCTD@cell_type_info$info[[3]]
S <- matrix(0, nrow(myRCTD@results$results_df), n)
rownames(S) <- rownames(myRCTD@results$results_df)
colnames(S) <- myRCTD@cell_type_info$info[[2]]

SCALE <- 10
for (i in seq_len(nrow(S))) {
  sc <- myRCTD@results$singlet_scores[[i]]
  S[i, names(sc)] <- exp((min(sc) - sc) / SCALE)
}
```

Quick visual check (using RCTD palette and geoms):

```{r}
plot_df <- cbind(myRCTD@spatialRNA@coords, S)
ggplot(plot_df, aes(x, y, color = Granule)) +
  geom_point(size = 0.6, alpha = 0.9) +
  theme_classic()
```

Focus subset (here we keep all spots):

```{r}
barcodes <- rownames(S)
S_mat <- S[barcodes, , drop=FALSE]
```

# Spatial graph $A$ (Delaunay)

We build a planar graph using the **spatialTV** package's `build_spatial_graph()` function. It creates a signed incidence matrix $A$ from a Delaunay triangulation: one row per edge with $\{+1,-1\}$ at the two incident spots.

```{r build-graph}
# Extract spatial coordinates
coords <- myRCTD@spatialRNA@coords[barcodes, ]

# Build spatial graph using spatialTV package function
A <- build_spatial_graph(coords, method = "delaunay")

# Check dimensions
cat("Graph dimensions:\n")
cat("  Number of spots (N):", ncol(A), "\n")
cat("  Number of edges (m):", nrow(A), "\n")
cat("  Density:", sprintf("%.2f%%", 100 * sum(A != 0) / length(A)), "\n")
```

# Initialize Dictionary $B$

The dictionary matrix $B$ defines mixture components. We use **spatialTV's** `initialize_dictionary()` which performs k-means clustering on $S$ to get data-driven initial components.

```{r initialize-B}
# Number of mixture components
D <- 3

# Initialize B using k-means on S
B <- initialize_dictionary(S_mat, D = D, method = "kmeans", seed = 42)

# Check constraints
cat("Dictionary dimensions: K x D =", paste(dim(B), collapse = " x "), "\n")
cat("Min value (should be >= eps):", sprintf("%.2e", min(B)), "\n")
```

# Fit Spatial Mixture Model

Now we run the main optimization using `fit_spatial_mixture()`. This alternates between optimizing mixture proportions $X$ and dictionary $B$ with spatial Total Variation regularization.

```{r fit-model}
# Set regularization parameter
lambda <- 1

# Fit the model
result <- fit_spatial_mixture(
  S = S_mat,
  A = A,
  B = B,
  lambda = lambda,
  it_X = 5000,
  it_B = 1000,
  log_every_X = 0,    # Set to 0 to suppress iteration logs
  log_every_B = 0
)

# Print results summary
print(result)
```

# Visualize Results

## Mixture Proportions

Visualize the optimized mixture components across space using the enhanced plotting function:

```{r plot-mixtures, fig.width=10, fig.height=4}
# Extract mixture matrix X
X <- result$X

# Use the enhanced plotting function from spatialTV
plot_components_spatial(
  X = X,
  coords = coords,
  point_size = 0.8,
  color_palette = "magma",
  ncol = 3,
  alpha = 0.9,
  high_res = TRUE,
  title_prefix = "Component"
)
```

For publication-quality figures, you can also add contour lines and adjust styling:

```{r plot-mixtures-enhanced, fig.width=10, fig.height=4, eval=FALSE}
# Enhanced version with contours
plot_components_spatial(
  X = X,
  coords = coords,
  point_size = 1.2,
  color_palette = "magma",
  ncol = 3,
  alpha = 0.95,
  add_contours = TRUE,
  contour_bins = 12,
  high_res = TRUE
)
```

## Dominant Component

Show which component is dominant at each spot:

```{r plot-dominant}
# Get dominant component
dominant_component <- apply(X, 1, which.max)

plot_df$Dominant <- factor(dominant_component)

ggplot(plot_df, aes(x = x, y = y, color = Dominant)) +
  geom_point(size = 0.6, alpha = 0.9) +
  scale_color_viridis(discrete = TRUE, option = "plasma") +
  theme_classic() +
  theme(aspect.ratio = 1) +
  labs(title = "Dominant Mixture Component per Spot",
       color = "Component")
```

## Cell Type Composition

Convert mixture representation back to cell type proportions and visualize:

```{r plot-celltypes}
# Convert to cell type proportions
cell_type_props <- components_to_celltypes(result$X, result$B)

# Get dominant cell type per spot
cell_type_names <- colnames(S_mat)
dominant_celltype <- get_dominant_celltype(
  result$X,
  result$B,
  cell_type_names = cell_type_names
)

# Plot dominant cell type
plot_df$CellType <- dominant_celltype

ggplot(plot_df, aes(x = x, y = y, color = CellType)) +
  geom_point(size = 0.6, alpha = 0.9) +
  theme_classic() +
  theme(
    aspect.ratio = 1,
    legend.position = "right"
  ) +
  labs(title = "Dominant Cell Type per Spot (from spatialTV)")
```

## Optimized Dictionary

Examine the learned dictionary to see which cell types contribute to each component:

```{r plot-dictionary, fig.height=5}
# Prepare dictionary for plotting
B_df <- as.data.frame(result$B)
colnames(B_df) <- paste0("Component_", 1:D)
B_df$CellType <- cell_type_names

# Reshape for plotting
B_long <- pivot_longer(
  B_df,
  cols = starts_with("Component"),
  names_to = "Component",
  values_to = "Weight"
)

# Plot
ggplot(B_long, aes(x = CellType, y = Weight, fill = Component)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Component, nrow = 3) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_blank()
  ) +
  labs(
    title = "Learned Dictionary: Cell Type Contributions to Each Component",
    y = "Weight",
    x = "Cell Type"
  )
```

## Saving High-Quality Figures

Export publication-ready figures:

```{r save-figures, eval=FALSE}
# Save component plots as PDF (vector graphics)
p <- plot_components_spatial(
  X = result$X,
  coords = coords,
  point_size = 1.5,
  color_palette = "magma",
  ncol = 3,
  alpha = 0.95,
  high_res = TRUE
)

ggsave("mixture_components.pdf", p, width = 12, height = 4, dpi = 300)

# Or save as high-resolution PNG
ggsave("mixture_components.png", p, width = 12, height = 4, dpi = 600)
```

# Model Diagnostics

Check objective function components:

```{r diagnostics}
cat("Model Fit Summary:\n")
cat("==================\n")
cat("Objective value:", sprintf("%.6e", result$obj), "\n")
cat("  - NLL component:", sprintf("%.6e", result$nll), "\n")
cat("  - TV penalty:", sprintf("%.6e", result$tv), "\n")
cat("  - Ratio (TV/NLL):", sprintf("%.3f", result$tv / result$nll), "\n\n")

cat("Optimization Parameters:\n")
cat("  Lambda (TV weight):", lambda, "\n")
cat("  X step size:", sprintf("%.2e", result$stepX), "\n")
cat("  B step size:", sprintf("%.2e", result$stepB), "\n")
```

# Summary

This demo showcased the complete **spatialTV** workflow:

1. ✅ Extracted cell type scores $S$ from RCTD deconvolution
2. ✅ Built spatial graph $A$ using `build_spatial_graph()`
3. ✅ Initialized dictionary $B$ using `initialize_dictionary()`
4. ✅ Optimized with `fit_spatial_mixture()` using TV regularization
5. ✅ Visualized mixture components and cell type compositions

## Key Features

- **Spatial coherence**: TV regularization enforces smooth transitions between neighboring spots
- **Low-rank representation**: Dictionary learning reduces noise and finds interpretable patterns
- **Data-driven**: K-means initialization adapts to your specific tissue architecture
- **Fast optimization**: C++ implementation with FISTA algorithm

## Next Steps

To further explore your results:

- **Adjust `lambda`** to control spatial smoothness vs data fit
- **Try different `D`** (number of components) to capture tissue complexity
- **Use `fit_spatial_mixture_adaptive()`** for adaptive edge weights
- **Explore color palettes** in `plot_components_spatial()`: "viridis", "magma", "plasma", "inferno", "rocket", "mako"
- **Add contours** with `add_contours = TRUE` to visualize spatial gradients
- **Export high-quality figures** using `ggsave()` for publications

For more information, visit: https://github.com/liutom445/spatialTV

