---
title: "Introduction to spatialTV"
author: "Tom Liu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spatialTV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `spatialTV` package implements spatially-aware cell type deconvolution for spatial transcriptomics data using Total Variation (TV) regularization. It combines:

1. **Low-rank factorization**: Learns a compact dictionary of cell type patterns
2. **Spatial regularization**: Enforces smoothness via TV penalty on spatial graphs
3. **Efficient optimization**: Uses FISTA algorithm with accelerated gradient descent
4. **Adaptive weighting**: Optionally learns edge-specific importance weights

## Installation

```{r, eval=FALSE}
# From CRAN (once published)
install.packages("spatialTV")

# Development version from GitHub
# install.packages("devtools")
devtools::install_github("liutom445/spatialTV")
```

## Quick Start

```{r quickstart, eval=FALSE}
library(spatialTV)

# Load your data
# S: cell type scores from RCTD or similar (N x K matrix)
# coords: spatial coordinates (N x 2 matrix)

# Build spatial graph
A <- build_spatial_graph(coords)

# Initialize dictionary
B_init <- initialize_dictionary(S, D = 3, method = "kmeans")

# Fit model
result <- fit_spatial_mixture(
  S = S,
  A = A,
  B = B_init,
  lambda = 2.0
)

# Extract results
X <- result$X  # Mixture proportions (N x D)
B <- result$B  # Cell type dictionary (K x D)

# Convert to cell type proportions
cell_props <- components_to_celltypes(X, B)
```

## Mathematical Framework

### Problem Formulation

Given:
- **S**: Cell type scores from deconvolution (N spots × K cell types)
- **A**: Spatial graph incidence matrix (m edges × N spots)
- **w**: Edge weights (m-dimensional)

We learn:
- **X**: Per-spot mixture components (N × D), rows on simplex
- **B**: Cell type dictionary (K × D), columns on simplex

The objective function is:

$$
\min_{X,B} \underbrace{-\sum_{i=1}^N \log z_i}_{\text{Data fidelity}} + 
\lambda \underbrace{\sum_{e=1}^m w_e \sum_{d=1}^D \sqrt{(AX)_{ed}^2 + \delta}}_{\text{Spatial TV penalty}}
$$

where the per-spot evidence is:
$$
z_i = \sum_{k=1}^K S_{ik} \sum_{d=1}^D X_{id} B_{kd} = \sum_k S_{ik}(XB^\top)_{ik}
$$

### Optimization Algorithm

The package uses an alternating FISTA approach:

1. **X-step**: Optimize mixture proportions with fixed B
   - Compute gradients of data term and TV penalty
   - FISTA update with momentum and simplex projection
   
2. **B-step**: Optimize dictionary with fixed X
   - Only data gradient (no regularization on B)
   - FISTA update with column-simplex projection

3. **w-step** (optional): Optimize edge weights via LP
   - Minimize weighted total variation
   - Subject to: $w_{\min} \leq w \leq w_{\max}$ and $|A|^\top w = \mathbf{1}$

## Detailed Example

### Data Preparation

```{r data_prep, eval=FALSE}
library(spatialTV)
library(ggplot2)

# Simulate example data or load real data
data(example_S_small)  # Cell type scores
data(example_A_small)  # Spatial graph

S <- example_S_small
A <- example_A_small

dim(S)  # 100 spots × 19 cell types
dim(A)  # ~300 edges × 100 spots
```

### Building the Spatial Graph

For real data with coordinates:

```{r spatial_graph, eval=FALSE}
# Assuming you have coordinates
coords <- cbind(x = runif(100, 0, 10), y = runif(100, 0, 10))

# Build Delaunay triangulation
A <- build_spatial_graph(coords, method = "delaunay")

# Check graph properties
cat("Number of edges:", nrow(A), "\n")
cat("Average degree:", 2 * nrow(A) / ncol(A), "\n")
```

### Dictionary Initialization

```{r init_dict, eval=FALSE}
# Method 1: K-means clustering (recommended)
B_kmeans <- initialize_dictionary(S, D = 3, method = "kmeans", seed = 123)

# Method 2: Random initialization
B_random <- initialize_dictionary(S, D = 3, method = "random", seed = 123)

# Method 3: Uniform initialization
B_uniform <- initialize_dictionary(S, D = 3, method = "uniform")

# Check constraints
colSums(B_kmeans)  # Should be all 1s
min(B_kmeans)       # Should be >= 1e-6
```

### Basic Optimization

```{r basic_opt, eval=FALSE}
# Fit model with fixed uniform weights
result <- fit_spatial_mixture(
  S = S,
  A = A,
  B = B_kmeans,
  lambda = 2.0,              # TV regularization strength
  delta = 1e-6,              # Soft-L1 smoothing
  it_X = 1500,               # Max iterations for X
  it_B = 500,                # Max iterations for B
  log_every_X = 250          # Logging frequency
)

# Print summary
print(result)

# Check objective value
cat("Objective:", result$obj, "\n")
cat("NLL:", result$nll, "\n")
cat("TV:", result$tv, "\n")
```

### Adaptive Weight Learning

```{r adaptive, eval=FALSE}
# Requires CVXR package
# install.packages("CVXR")

result_adaptive <- fit_spatial_mixture_adaptive(
  S = S,
  A = A,
  B_init = B_kmeans,
  lambda = 2.0,
  outer_iter = 5,            # Number of alternating iterations
  wmin = 0.01,               # Minimum edge weight
  wmax = 0.50,               # Maximum edge weight
  learn_weights = TRUE
)

# Check convergence
print(result_adaptive$history)

# Plot objective over iterations
plot(result_adaptive$history$iter, 
     result_adaptive$history$obj,
     type = "b", xlab = "Iteration", ylab = "Objective")

# Examine weight distribution
hist(result_adaptive$w, breaks = 30, 
     main = "Edge Weight Distribution",
     xlab = "Weight", col = "skyblue")
```

### Extracting Results

```{r extract_results, eval=FALSE}
# Get mixture components
X <- result$X  # N x D matrix

# Convert to cell type proportions
cell_props <- components_to_celltypes(X, result$B)
dim(cell_props)  # N x K

# Get dominant cell type per spot
dominant <- get_dominant_celltype(X, result$B, 
                                  cell_type_names = colnames(S))
table(dominant)

# Get top 3 cell types per spot
top_types <- get_top_celltypes(X, result$B, K = 3,
                               cell_type_names = colnames(S))
head(top_types$types)
head(top_types$proportions)
```

## Visualization

### Spatial Maps

```{r viz_spatial, eval=FALSE}
library(ggplot2)

# Prepare data frame
plot_df <- data.frame(
  x = coords[, 1],
  y = coords[, 2],
  celltype = dominant
)

# Plot dominant cell types
ggplot(plot_df, aes(x = x, y = y, color = celltype)) +
  geom_point(size = 2) +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Spatial Cell Type Map")

# Plot specific cell type proportion
plot_df$granule_prop <- cell_props[, "Granule"]

ggplot(plot_df, aes(x = x, y = y, color = granule_prop)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Granule Cell Proportion")
```

### Convergence Diagnostics

```{r viz_convergence, eval=FALSE}
# For adaptive model
library(ggplot2)
library(tidyr)

# Plot multiple metrics
hist_long <- pivot_longer(
  result_adaptive$history,
  cols = c(obj, nll, tv),
  names_to = "metric",
  values_to = "value"
)

ggplot(hist_long, aes(x = iter, y = value, color = metric)) +
  geom_line() +
  geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal() +
  labs(title = "Convergence Over Iterations")
```

## Choosing Hyperparameters

### Regularization Strength (lambda)

The `lambda` parameter controls the trade-off between data fidelity and spatial smoothness:

- **Small lambda** (e.g., 0.1-1): Less smoothing, more detailed patterns
- **Medium lambda** (e.g., 1-5): Balanced, recommended for most applications
- **Large lambda** (e.g., 5-20): Strong smoothing, may over-regularize

```{r tune_lambda, eval=FALSE}
lambda_values <- c(0.5, 1, 2, 5, 10)
results <- list()

for (lam in lambda_values) {
  results[[as.character(lam)]] <- fit_spatial_mixture(
    S = S, A = A, B = B_kmeans,
    lambda = lam,
    it_X = 1000,
    log_every_X = 0  # Suppress logging
  )
}

# Compare metrics
sapply(results, function(r) {
  c(obj = r$obj, nll = r$nll, tv = r$tv)
})
```

### Number of Components (D)

Choose D based on:
- **Biological knowledge**: Expected number of distinct microenvironments
- **Model selection**: Cross-validation or information criteria
- **Practical constraint**: Typically D = 3-8

```{r tune_D, eval=FALSE}
D_values <- c(2, 3, 4, 5)
results_D <- list()

for (D in D_values) {
  B_init <- initialize_dictionary(S, D = D, method = "kmeans")
  results_D[[as.character(D)]] <- fit_spatial_mixture(
    S = S, A = A, B = B_init,
    lambda = 2.0,
    log_every_X = 0
  )
}

# Compare objectives
sapply(results_D, `[[`, "obj")
```

## Performance Metrics

```{r metrics, eval=FALSE}
# Compute comprehensive metrics
metrics <- result_adaptive$final_metrics

# Interpretation:
# - neighbor_agreement: Fraction of edges with same dominant type (higher is smoother)
# - entropy: 0 = deterministic, 1 = uniform (lower is sharper)
# - sharpness: Mean of max proportion per spot (higher is more confident)

print(metrics)
```

## Tips for Production Use

1. **Start with uniform weights**: Use `fit_spatial_mixture()` first
2. **Use adaptive weights selectively**: Only when edges have different importance
3. **Monitor convergence**: Check that objective decreases monotonically
4. **Validate biologically**: Compare with known markers and spatial structure
5. **Consider computation**: Adaptive mode slower due to LP solve (~2-3x)

## Session Info

```{r sessioninfo}
sessionInfo()
```

## References

1. Beck, A. and Teboulle, M. (2009). "A Fast Iterative Shrinkage-Thresholding Algorithm for Linear Inverse Problems." *SIAM Journal on Imaging Sciences*, 2(1), 183-202.

2. Cable, D. M., et al. (2021). "Robust decomposition of cell type mixtures in spatial transcriptomics." *Nature Biotechnology*.

3. Rudin, L. I., Osher, S., & Fatemi, E. (1992). "Nonlinear total variation based noise removal algorithms." *Physica D: nonlinear phenomena*, 60(1-4), 259-268.
